Regularising Lotka--Volterra interaction matrices by species grouping: optimal groupings are unpredictable

### 2022-05-10 article.rnw

Models have long been used in community ecology \parencite{lotka1925elements}.
They allow us to make predictions about things we cannot directly observe, be it the response of a community to novel pressures \parencite{novak2016characterizing}, or the stability of an entirely novel assemblage \parencite{maynard2020predicting}.
The ability to make these predictions is becoming increasingly important as global change accelerates, introducing alien species and applying climatic or other anthropogenic pressures to communities worldwide \parencite{XXX}.

Making practical use of a model requires experimentally or observationally determining the values of each of the model's parameters.
This poses a particular problem for models that include species interactions due to the sheer number of parameters that must be determined.
Even the simplest models with no environmental dependence require one parameter to characterise the interaction of each pair of species \parencite{novak2016characterizing, XXX}.
Hence the number of parameters scales quadratically with the number of species; at least \num{100} parameters are required for the interactions in a community of ``just'' \num{10} species.
Determining these parameters experimentally requires at least one experiment per parameter, and preferably more \parencite{hart2018quantify}.
This is expensive, and requires a lot of effort.
As such, model-based predictions of real communities are typically restricted to a small number of focal species \parencite{ovaskainen2017species, XXX}.
However, real communities can contain hundreds of species \parencite{cornell1999unsaturation}, or even thousands for microbial communities assessed by genetic metabarcoding \parencite{warton2015variables}.
To bridge this gap between the size of real communities and our ability to parameterise them, we need an ecologically-defensible way to reduce the number of parameters that predictive models require.

One simplifying approach is to assume that most pairs of species do not interact or interact weakly, and hence setting many of the parameters to zero \parencite{mutshinda2009drives}.
However, weak interactions have been shown to be very important to the stability of ecological communities \parencite{wootton2016many}, and setting these to zero may lose important information.
Another approach assume that species fall into groups, such that all species in the same group interact with species in other groups in an identical fashion, thereby collapsing pairwise interaction parameters for all the species in that group into sets of common parameters \parencite{ovaskainen2017species,weiss2021disentangling}.

In ecological contexts, such groups have been defined \emph{a priori} by phylogeny at some taxonomic level or by functional groupings \parencite{uriarte2004spatially,ovaskainen2017species,weiss2021disentangling}.
The argument for grouping based on phylogeny relies on the principle that closely related species will fill similar ecological niches, and hence will interact with other species in a similar fashion \parencite{godoy2014phylogenetic}.
However, empirical tests of this principle have found remarkably little support for it \parencite{godoy2014phylogenetic,cahill2008does,mayfield2010opposing}.
On the other hand, functional groupings have also proven to be poor predictors of species interactions \parencite{bret2008plant}.
While attempts have been made to define functional groupings that are more useful in a modelling context \parencite{boulangeat2012improving}, these rely upon measurements of biotic interactions to define such groups.
As such, integrating these into species interaction models leaves us with a circular argument: a functional group is a group of species that interact with others in a similar way, and we therefore assume that all the species in a functional group interact with others in a similar way.
In all, we are left with the question of how best to group species in community models, and whether we can do so in a meaningful way using less data than required for a fully pairwise parameterisation.

Additionally, studies that use grouping to simplify models of species interactions have always made one untested assumption: that the same grouping is equally valid for how species affect others as for how species respond to others \parencite{XXX}.
There is good reason to question this assumption, as it is known that different traits can govern species' competitive effects than govern their responses to competition \parencite{goldberg1991competitive}.
As such, the best groupings of species may be different for each of the effects and responses.

We propose that species might be best partitioned into groups without regard to any \emph{a priori} definition but instead based purely upon which grouping gives the most parsimonious model.
% DBS: We can say something here about how this data-informed approach has worked very well for other clustering problems, such as modularity in networks. I can point to some very good references along those lines.
In this study, we investigate the usefulness and correctness of such an approach using experimental \emph{Drosophila} fecundity data provided by \textcite{terry2021natural}.
We fit the Lotka--Volterra model, with grouped species, to find the grouping that gives the most parsimonious model.
We group species independently with regard to their effects upon other species and their responses to other species, to find whether the appropriate groupings might differ for these two cases.
%TODO: Mention the comparison to traits (body size)?

@@

To assess the parsimony of the different species groupings, we exhaustively fit all \num{\Sexpr{numbers::bell(6)^2}} possible models.
As a measure of parsimony, we selected AIC over BIC on the basis that we want to find the best model for making ecological predictions, rather than seeking the ``true'' model \parencite{aho2014model}.
Further, AIC is more conservative in promoting grouping, as BIC penalises number of parameters more for any reasonably sized data set.
Some of the models had a relatively high number of parameters compared to the number of observations (\num{\Sexpr{num_obs}} observations; up to \num{\Sexpr{num_species + num_species^2}} parameters), so we used AICc to correct for this.
As such, we calculated the {\ic} for each model.

{\ic} values can be compared to determine which model is the most parsimonious, but cannot directly be interpreted to find the significance of the difference in parsimony.
As such, we also converted the set of {\ic} values to Akaike weights \parencite{wagenmakers2004aic}.
These give the probability that each model is the best among the models assessed.

From the Akaike weights, we calculated two important statistics.
Firstly, the total Akaike weight of models with a lower {\ic} than the entirely ungrouped model; this is the probability that some model is more parsimonious than the ungrouped model.
Secondly, the total Akaike weight of models that used the same grouping for the rows as for the columns; this is the probability that the most parsimonious model grouped species equivalently in their competitive effects as in their competitive responses.
To evaluate the significance of the latter, we performed a Monte Carlo permutation test, comparing the total Akaike weight of those models that grouped rows and columns equivalently to the total Akaike weight of a random equally-sized subset of the models.
%This is a test of whether the most parsimonious model is more likely to group the species equivalently in the rows as in the columns than it would be by chance.

@@

	\caption{%
		A histogram of the {\ic} values of species interaction models with all possible groupings of interaction coefficients.
		The green line shows the {\ic} value (\icnum{\Sexpr{brute_data$fully_separated_statistics[[ic]]}}) of the ungrouped model, in which all competition coefficients are fitted independently.
		The red line shows the {\ic} value (\icnum{\Sexpr{brute_data$fully_grouped_statistics[[ic]]}}) of the fully grouped model, where every pair of species shares a single interaction coefficient.
	}
	\label{fig:ic}
\end{figure}

The ``best'' model, as indicated by {\ic}, was not the best with any surety; its Akaike weight indicated a probability of only \probnum{\Sexpr{max(brute_data$statistics[[ic_weight]])}} that it was the best.
\Cref{fig:best-model} shows the groupings and alpha values of this ``best'' model.
The total Akaike weight assigned to models more parsimonious than the ungrouped model was \probnum{\Sexpr{sum(brute_data$statistics[[ic_weight]][brute_data$statistics[[ic]] < brute_data$fully_separated_statistics[[ic]]])}}; this is the probability that some grouped model was more parsimonious than the ungrouped model.

Considering only models in which the rows and columns were grouped using the same grouping, the {\ic} values ranged from \icnum{\Sexpr{min(brute_data$equivalently_grouped_statistics[[ic]])}} to \icnum{\Sexpr{max(brute_data$equivalently_grouped_statistics[[ic]])}}; neither the best nor worst overall models had equivalently grouped rows and columns.
The total Akaike weight assigned to these models was \probnum{\Sexpr{brute_data$equivalently_grouped_weight}}; more than expected by chance (\probnum{\Sexpr{brute_data$equivalently_grouped_expected_weight}}), but not significantly so (Monte Carlo permutation test, $p=\probnum{\Sexpr{brute_data$equivalently_grouped_weight_p #$}}$).
This indicates that the most parsimonious model is not especially likely to be one that groups the rows and columns equivalently.

@@

\begin{figure}
	\centering
	\begin{subfigure}{\textwidth}
		\centering
		<<echo=FALSE>>=
			weighted_coclassification_kable(
				weighted_col_coclassification_matrix[species_output_order,species_output_order],
				colourmap="coclassification", diagonal_colour="coclassification diagonal"
			)
		@
		\caption{Weighted coclassification of competitive effects (columns of the interaction matrix).}
		\label{fig:coclassification-cols}
	\end{subfigure}

@@

We have shown that it is possible to fit a more parsimonious model of species interactions by grouping the interaction coefficients of species \emph{a posteriori}, and hence reducing the number of free parameters in the model.
Indeed, the results show that grouping species was likely to give a more parsimonious model (probability \probnum{\Sexpr{sum(brute_data$statistics[[ic_weight]][brute_data$statistics[[ic]] < brute_data$fully_separated_statistics[[ic]]])}}) than leaving them ungrouped.
Which grouping would give the most parsimonious model, however, could not be determined with confidence.
This result stands in contrast to the popular wisdom in ecology that species should form some sort of clear groups, be they the functional groups so often used \parencite{blondel2003guilds}, or taxonomic groups per a relationship between phylogeny and ecological niche \parencite{godoy2014phylogenetic}.
It also contrasts with previous studies that have succeeded in finding clear groups, such as \textcite{allesina2009food}, who used AIC to find the best grouping of species, but in a binary food web rather than a quantitative competitive context.
Perhaps species simply do not group to the same extent in a competitive context.
The results suggest that this is unlikely, as grouping was still likely to increase parsimony.
Further, there is no reason to suspect that species would not have clear similarities or dissimilarities in their ecological niches, and therefore their competitive effects and responses.
Perhaps the lack of clarity in the groupings is instead a product of the study system chosen.

All the species in this study were members of the same genus, \emph{Drosophila}, and as such were all rather similar.
A study system with species spanning multiple functional groups, or a broader taxonomy, might have shown clearer groupings.
Alternatively, the small number of species in the study might not have been sufficient for clear groups to be defined.
Alas, the brute force method we employed in this study has a very poor algorithmic complexity, preventing the application of the method to studies with more than seven or eight species.
As such, a better method would be required in order to test for grouping of species interactions among a larger set of species.

The intransitivity of coclassification in some instances even suggests that there may not be a single ``best'' grouping of the species, but rather two or more quite distinct and mutually exclusive groupings of similar validity.
This could suggest a Bayesian approach, in which the outcome is not a definitive grouping, but a distribution of valid groupings.
Alternatively, it may suggest a mixed membership approach in which a species can be a partial part of more than one group \parencite{airoldi2008mixed}, reminiscent of the community-level drivers of \textcite{ovaskainen2017species}.

@@

All in all, the method of \emph{a posteriori} parsimony-based selection of groups has not proven to be greatly effective, at least on such a closely related set of species.
However, the results suggest some advice which might be borne in mind whenever selecting groupings \emph{a priori} for use in species interaction models.
Firstly, there may be no single best grouping, but perhaps multiple rather distinct groupings which each perform similarly.
Furthermore, one should not implicitly assume that species are grouped in the same way with regard to their effects upon other species and their response to them; distinct groupings of the two might well be considered.

>>>>

Models have long been used in community ecology \parencite{lotka1925elements}.
They allow us to make predictions about things we cannot directly observe, be it the response of a community to novel pressures \parencite{novak2016characterizing} or the stability of an entirely novel assemblage \parencite{maynard2020predicting}.
The ability to make these predictions is becoming increasingly important as global change accelerates, introducing alien species and applying climatic or other anthropogenic pressures to communities worldwide \parencite{XXX}.

Making practical use of a model requires experimentally or observationally determining the values of each of the model's parameters.
This poses a particular problem for models that include species interactions due to the sheer number of parameters that must be determined.
Even the simplest models with no environmental dependence require one parameter to characterise the interaction of each pair of species \parencite{novak2016characterizing, XXX}.
% DBS: Hmm. Should the above two or rephrased to embrace the directionality of what's going on?
Hence the number of parameters scales quadratically with the number of species; at least \num{100} parameters are required for the interactions in a community of ``just'' \num{10} species.
Determining these parameters experimentally requires at least one experiment per parameter, and preferably more \parencite{hart2018quantify}.
This is understandably expensive and requires substantial effort.
As such, model-based predictions of real communities are typically restricted to a small number of focal species \parencite{ovaskainen2017species, XXX}.
However, real communities can contain hundreds of species \parencite{cornell1999unsaturation}, or even thousands for microbial communities assessed by genetic metabarcoding \parencite{warton2015variables}.
To bridge this gap between the size of real communities and our ability to parameterise models for them, we need an ecologically-defensible way to reduce the number of parameters that predictive models require.

One simplifying approach is to assume that most pairs of species do not interact or interact weakly, and hence setting many of the parameters to zero \parencite{mutshinda2009drives}.
However, weak interactions have been shown to be very important to the stability of ecological communities \parencite{XXX, wootton2016many}, and setting these to zero may lose important information.
Another approach assumes that species fall into groups, such that all species in the same group interact with species in other groups in an identical fashion, thereby collapsing pairwise interaction parameters for all the species in that group into sets of common parameters \parencite{ovaskainen2017species,weiss2021disentangling}.

In ecological contexts, such groups have been defined \emph{a priori} by phylogeny at some taxonomic level or by functional similarity \parencite{uriarte2004spatially,ovaskainen2017species,weiss2021disentangling}.
% please add a reference to http://dx.doi.org/10.1086/713082 to the above
The argument for grouping based on phylogeny relies on the principle that closely related species will fill similar ecological niches, and hence will interact with other species in a similar fashion \parencite{XXX,godoy2014phylogenetic}.
However, empirical tests of this principle have found remarkably little support for it \parencite{godoy2014phylogenetic,cahill2008does,mayfield2010opposing}.
On the other hand, functional groupings have also proven to be poor predictors of species interactions \parencite{XXX, bret2008plant}.
While attempts have been made to define functional groupings that are more useful in a modelling context \parencite{boulangeat2012improving}, these rely upon measurements of biotic interactions to define such groups.
As such, integrating these into species interaction models leaves us with a circular argument: a functional group is a group of species that interact with others in a similar way, and we therefore assume that all the species in a functional group interact with others in a similar way.
In all, we are left with the question of how best to group species in community models, and whether we can do so in a meaningful way using less data than required for a fully pairwise parameterisation.

To the best of our knowledge, studies that use grouping to simplify models of species interactions have also made one additional, untested assumption: that the same grouping is equally valid for how species affect others as for how species respond to others \parencite{XXX}.
There is good reason to question this assumption, as it is known that different traits can govern species' competitive effects than govern their responses to competition \parencite{goldberg1991competitive}.
% DBS: Maybe it's worth adding a concrete example here for the more empirical reader/reviewer.
As such, the best groupings of species may be different for each of the effects and responses.

We propose that species might be best partitioned into groups without regard to any \emph{a priori} definition but instead based purely upon which grouping gives the most parsimonious model.
% DBS: We can say something here about how this data-informed approach has worked very well for other clustering problems, such as modularity in networks. I can point to some very good references along those lines.
In this study, we investigate the usefulness and correctness of such an approach using experimental data from competition between different \emph{Drosophila} species provided by \textcite{terry2021natural}.
We fit a Lotka--Volterra model to these data to find the groupings that give the most parsimonious model.
While doing so, we group species independently with regard to their effects upon other species and their responses to other species, to find whether the data-informed groupings differ between these two cases.
%TODO: Mention the comparison to traits (body size)?
% DBS: If we leave in the traits, then we should mention them at least briefly.

@@

To assess the parsimony of the different species groupings, we exhaustively fit all \num{\Sexpr{numbers::bell(6)^2}} possible models and computed the resulting {\ic} for those fits.
As a measure of parsimony, we selected AICc over BIC on the basis that we want to find the best model for making ecological predictions rather than seeking the ``true'' model \parencite{aho2014model}.
Furthermore, AICc is more conservative in promoting grouping as BIC imposes a greater penalty for the number of parameters for any reasonably sized data set.
Some of the models had a relatively high number of parameters compared to the number of observations (\num{\Sexpr{num_obs}} observations compared to up to \num{\Sexpr{num_species + num_species^2}} parameters), so we used AICc to correct for this ``small sample size'' effect.
{\ic} values can be compared to determine which model is the most parsimonious, but they cannot directly be interpreted to find the significance of the difference in parsimony.
As such, we also converted the set of {\ic} values to Akaike weights \parencite{wagenmakers2004aic}.
These give the probability that each model is the best among the models assessed.

Using these Akaike weights, we calculated two important statistics.
First, the total Akaike weight of models with a lower {\ic} than the entirely ungrouped model; this is the probability that some model is more parsimonious than the fully ungrouped model.
Second, the total Akaike weight of models that used the same grouping for the rows as for the columns; this is the probability that the most parsimonious model grouped species equivalently in their competitive effects as in their competitive responses.
To evaluate the significance of the latter, we performed a Monte Carlo permutation test, comparing the total Akaike weight of those models that grouped rows and columns equivalently to the total Akaike weight of a equally sized random subset of the models.
%This is a test of whether the most parsimonious model is more likely to group the species equivalently in the rows as in the columns than it would be by chance.

@@

	\caption{%
		Histogram of the {\ic} values of Lotka--Volterra model fit with all possible groupings of interaction coefficients.
		The green line shows the {\ic} value (\icnum{\Sexpr{brute_data$fully_separated_statistics[[ic]]}}) of the ungrouped model, in which all competition coefficients are fitted independently.
		The red line shows the {\ic} value (\icnum{\Sexpr{brute_data$fully_grouped_statistics[[ic]]}}) of the fully grouped model, where every pair of species shares a single interaction coefficient.
	}
	\label{fig:ic}
\end{figure}

The ``best'' model, as indicated by {\ic}, was not the best with any surety; its Akaike weight indicated a probability of only \probnum{\Sexpr{max(brute_data$statistics[[ic_weight]])}} that it was the best.
\Cref{fig:best-model} shows the groupings and alpha values of this ``best'' model.
Despite it being just over $7$ {\ic} units from the most parsimonious model, the total Akaike weight assigned to models more parsimonious than the ungrouped model was \probnum{\Sexpr{sum(brute_data$statistics[[ic_weight]][brute_data$statistics[[ic]] < brute_data$fully_separated_statistics[[ic]]])}}; this is the probability that some grouped model was more parsimonious than the ungrouped model.

Considering only models in which the rows and columns were grouped using the same grouping, the {\ic} values ranged from \icnum{\Sexpr{min(brute_data$equivalently_grouped_statistics[[ic]])}} to \icnum{\Sexpr{max(brute_data$equivalently_grouped_statistics[[ic]])}}; neither the best nor worst overall models had equivalently grouped rows and columns.
The total Akaike weight assigned to these models was \probnum{\Sexpr{brute_data$equivalently_grouped_weight}}; this was more than expected by chance (\probnum{\Sexpr{brute_data$equivalently_grouped_expected_weight}}), but not significantly so (Monte Carlo permutation test, $p=\probnum{\Sexpr{brute_data$equivalently_grouped_weight_p #$}}$).
This indicates that the most parsimonious model is not especially likely to be one that groups the rows and columns equivalently.

@@

% DBS: Most journals frown upon or do not allow overly verbose subtitles on figures. So I've cut them down to be a bit more concise.
% DBS: But the fact that the labels say effects and responses reminds me of a lingering "doubt" I had when reading the main text. We talk almost entirely about rows and columns but there are a number of junctures where I feel like it would be easier to follow for people who aren't us if we were talking about responses and effects. As long as we're consistent, I'm fine either way. But we need to remember we're writing the paper for the audience and not for ourselves. ;)
\begin{figure}
	\centering
	\begin{subfigure}{\textwidth}
		\centering
		<<echo=FALSE>>=
			weighted_coclassification_kable(
				weighted_col_coclassification_matrix[species_output_order,species_output_order],
				colourmap="coclassification", diagonal_colour="coclassification diagonal"
			)
		@
		\caption{Coclassification of competitive effects}
		\label{fig:coclassification-cols}
	\end{subfigure}

@@

We have shown that it is possible to fit a more parsimonious model of species interactions by grouping the interaction coefficients of species \emph{a posteriori}, and hence reducing the number of free parameters in the model.
Indeed, the results show that grouping species was likely to give a more parsimonious model (probability \probnum{\Sexpr{sum(brute_data$statistics[[ic_weight]][brute_data$statistics[[ic]] < brute_data$fully_separated_statistics[[ic]]])}}) than leaving them ungrouped.
Which grouping would give the most parsimonious model, however, could not be determined with confidence.
This result stands in contrast to the popular wisdom in ecology that species should form some sort of clear groups, be they the functional groups so often used \parencite{XXX, blondel2003guilds}, or taxonomic groups per a relationship between phylogeny and ecological niche \parencite{XXX, godoy2014phylogenetic}.
It also contrasts with previous studies that have succeeded in finding clear groups, such as \textcite{allesina2009food}, who used AIC to find the best grouping of species, but in a binary food web rather than a quantitative competitive context.
% DBS: We're probably doing a disservice to the "groups in ecological communities" literature by only citing the Allesina paper above.

One possible explanation is that species simply do not group to the same extent in a competitive context.
That said, the results suggest that this is unlikely as grouping was still likely to increase parsimony.
Further, there is no reason to suspect that species would not have clear similarities or dissimilarities in their ecological niches, and therefore their competitive effects and responses \parencite{XXX}.
In the present case, we cannot rule out the possibility that a lack of clarity in the groupings is instead a product of the study system.
All the species in this study were members of the same genus, \emph{Drosophila}, and as such were all rather similar.
A study system with species spanning multiple functional groups, or a broader taxonomy, might have shown clearer groupings.
Alternatively, the small number of species in the study might not have been sufficient for clear groups to be defined.
Alas, the brute force method we employed in this study has a very poor algorithmic complexity, preventing the application of the method to studies with more than seven or eight species. %DBS: Add in the number of models for both of these to emphasize the point.
As such, a better method would be required in order to test for grouping of species interactions among a larger set of species.

% DBS: I think we should swap the order of the following two paragraphs.

The intransitivity of coclassification in some instances even suggests that there may not be a single ``best'' grouping of the species, but rather two or more quite distinct and mutually exclusive groupings of similar validity. A single trait like body size might therefore not be able to capture the multivariate nature of species' ecological niches \parencite{XXX}.
This could suggest a Bayesian approach is worth exploring in the future, in which the outcome is not a definitive grouping but a distribution of valid groupings.
Alternatively, it may suggest adopting a mixed-membership approach in which a species can belong to more than one group \parencite{airoldi2008mixed}, reminiscent of the community-level drivers of \textcite{ovaskainen2017species}.

@@

% DBS: Never shoot yourself in the foot in your concluding paragraph! The reviewers will already be primed to do this for you, so I always think it's better to take a glass half full perspective.
% DBS: Also, there is no hard and fast rule to say we need a "sum it all up" concluding paragraph. Each of the statements below is pretty much already stated above but with more depth. I think we could actually do away with it and lose very little.
All in all, the method of \emph{a posteriori} parsimony-based selection of groups has not proven to be greatly effective, at least on such a closely related set of species.
However, the results suggest some advice which might be borne in mind whenever selecting groupings \emph{a priori} for use in species interaction models.
Firstly, there may be no single best grouping, but perhaps multiple rather distinct groupings which each perform similarly.
Furthermore, one should not implicitly assume that species are grouped in the same way with regard to their effects upon other species and their response to them; distinct groupings of the two might well be considered.
